---
title: "The fiveSTAR Package"
author: "Rachel Marceau West"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The *fiveSTAR* package implements the 5-step Stratified Testing and Amalgamation Routine (5-STAR) algorithm for structured patient segregation and testing of overall treatment effect in a randomized clinical trial within a heterogeneous patient population. 

Given response (yy), treatment arm information (arm), and data frame of all candidate covariates (X), the fiveSTAR package runs the 5-STAR algorithm, giving homogeneous risk strata, final amalgamated treatment effect and corresponding p-value.

The 5-STAR algorithm consists of 5 key steps:

1. Pre-specify baseline covariates that may be prognostic for survival
2. Remove `noise' covariates from the pre-specified candidate list
3. Segregate the overall population into subpopulations (risk strata) of pronostically homogeneous patients
4. Perform treatment comparisons within each formed risk stratum
5. Amalgamate results across formed risk strata for overall hypothesis test and estimation

Full details of the algorithm can be found in the manuscript: 

Mehrotra D.V., Marceau West R. 2020. Survival Analysis Using a 5-Step Stratified Testing and Amalgamation Routine (5-STAR) in Randomized Clinical Trials. Statistics in Medicine  39(30):4724-4744.

## Quick Start

To demonstrate use of the package, we consider an example survival dataset from 
a hypothetical randomized clinical trial consisting of 600 patients, where 300 
are randomly assigned to test treatment and the remaining 300 are assigned to 
the control. 

Our example data set consists of 53 columns: ***time***, the observed survival time,
***status***, the censoring indicator (1 indicates an observed event and 0 indicates 
censoring), ***arm***, the assigned treatment arm (1 indicates test and 0 indicates 
control), and 50 candidate covariates: ***X1-X25*** are binary factor variables 
and ***X26-X50*** are continuous covariates.

```{r loaddata}
library(fiveSTAR)
data(survdata, package="fiveSTAR")
time = survdata$time
status = survdata$status
arm = survdata$arm
X = survdata[,!(colnames(survdata) %in% c("time","status","arm"))]
```

```{r displaydata, echo=FALSE}
library(DT)
DT::datatable(survdata,options=list(scrollX=13))
```

<!-- A standard Cox PH model applied to this data fails to detect any treatment benefit from the new test treatment, with estimated hazard ratio and corresponding 95% confidence interval of 0.87 (0.70, 1.08) and one sided logrank p-value of 0.105. -->

<!-- ```{r coxph} -->
<!-- library(survival) -->
<!-- coxfit = coxph(Surv(time,status) ~ arm) -->
<!-- summary(coxfit) -->
<!-- ``` -->

<!-- However, the smoothed log hazard curves show evidence of deviation from the proportional hazards assumption, in agreement with the Grambsch-Therneau (1994) test of PH -->

<!-- ```{r nonPH} -->
<!-- library(survminer) -->
<!-- KMfit = survfit(Surv(time,status)~arm,data=survdata) -->
<!-- KMplot = ggsurvplot(KMfit, conf.int=FALSE, legend.labs = c("Control","Test"), -->
<!--                       xlab="Time (Months)", ylab="Survival", -->
<!--                       legend.title="Treatment", -->
<!--                       risk.table=TRUE,break.time.by=3,data=survdata) -->
<!-- KMplot -->

<!-- library(bshazard) -->
<!-- bsA = bshazard(formula = Surv(time,status) ~ 1, data = survdata[arm == 1,]) -->
<!-- bsB = bshazard(formula = Surv(time,status) ~ 1, data = survdata[arm == 0,]) -->

<!-- #log-hazard plots.. -->
<!-- Aloghaz_overall = cbind(bsA$time,log(bsA$hazard),log(bsA$lower.ci), -->
<!--                         log(bsA$upper.ci),1) -->

<!-- Bloghaz_overall = cbind(bsB$time,log(bsB$hazard),log(bsB$lower.ci), -->
<!--                         log(bsB$upper.ci),0) -->

<!-- logmat_overall = rbind(Aloghaz_overall,Bloghaz_overall) -->
<!-- colnames(logmat_overall) = c("time","loghazard","lower.loghazard", -->
<!--                               "upper.loghazard","arm") -->
<!-- logmat_overall = as.data.frame(logmat_overall) -->

<!-- bshazoverall = ggplot(data=logmat_overall,aes( -->
<!--     x=time,y=loghazard,color=as.factor(arm),fill=as.factor(arm))) + -->
<!--     geom_line() + geom_ribbon(aes(x=time,ymin=lower.loghazard, -->
<!--                                   ymax=upper.loghazard),alpha=0.5) + -->
<!--     theme_classic() + xlab(paste0("Time (Months)")) + -->
<!--     ylab("Estimated log(Hazard)") + -->
<!--     scale_fill_discrete(name="Treatment",labels=c("Control","Test")) + -->
<!--     scale_color_discrete(name="Treatment",labels=c("Control","Test")) + -->
<!--     scale_x_continuous(breaks=seq(0,max(time),3)) -->

<!-- ``` -->

In truth, there is heterogeneity in subject prognosis. In particular, there are
four prognostic risk strata defined by three covariates (X1, X2, and X26 <= or > 0.4). 
In order to improve estimation via transparent patient risk segregation, we apply the 5-STAR algorithm to the data, inputting response yy, a right-censored survival object, arm, the treatment allocation information, and X, a matrix of candidate covariates specified in Step 1 of the algorithm. Note that X may contain numeric, character, factor, and ordered factor variables.

```{r fivestar, results='hide', fig.show='hide'}
library(survival)
cc = run5STAR(yy=Surv(time,status), arm=arm, X=X, measure="HR")
```

run5STAR() is the main worker function calling all other subfunctions. 
The workflow follows the described process:

1. Check/prepare input data -> prep5STAR()
2. Perform filtering -> filter5STAR()
3. Form trees -> fittrees()
4. Perform estimation within each formed strata -> coxbystrata() [HR] or maaftbystrata() [TR]
5. Aggregate -> minPadapHR()

### Data Cleaning and Preparation within 5-STAR

Data is prepared by first converting all character variables to factors. Next, 
proportion of missing data within each covariate is checked against the missingness
threshold as set by the missThreshold value in run5STAR(). missThreshold may be
a single number, in which case all covariates with proportion missingnes over the 
inputted threshold will be removed prior to analysis. It can also be a vector of 
length two, where the second number is a hard threshold as described above, and the first
is a softer threshold, removing covariates with missingness between the two values 
unless there is sufficient evidence they may have prognostic value, as evidenced by
a test of correlation between the blinded subject scores (e.g., logrank scores for
survival data) and the covariate. By default, missThreshold = c(0.1,0.2) is used,
indicating all covariates with >20% missingness will be removed and those with 10-20%
missingness will be removed unless there is sufficient evidence they may be prognostic 
(e.g., correlated with logrank scores). 

Covariates are also removed prior to analysis if they have no chance of splitting 
in the algorithm, e.g., if there are no minor categories of a factor variable with
the minimum number of subjects required per terminal node, as set by minbucket in the 
tree.hyper variable of run5STAR() (modified via tree_control()). Finally, factor covariates
with over 31 categories are removed prior to analysis as the Ctree algorithm is 
unable to handle such covariates. 

Details on removed covariates will be printed to the screen when verbose is set to 1 or higher. 
Further, a list of all covariates kept after data cleaning is given in the final output
of run5STAR(). For our example data, all covariates were clean and kept for the analysis.

```{r prep5STAR}
cc$inputcovs
```


### Filtering

As per the 5-STAR algorithm, the next step is filtering out any covariates with no 
evidence of assocation between response (e.g., survival times). The fiveSTAR 
package currently allows users to do this via Cox elastic net regression (default;
recommended for most scenarios) or random forest. The filtering method and related 
parameters are controlled through the filter.hyper option in run5STAR, which 
expects as input a list of options controlled through the filter_control() function. 

We consider the Cox Elastic Net filtering option (method='ENET'), which performs a 
grid search over mixing parameters and cross-validation within each mixing 
parameter to jointly select the optimal ENET parameters. By deafult, 10 cross 
valdidation folds are used to select the tuning parameter and a grid from 0.05 
to 0.95 with step size of 0.05 is used. These can be changed via the 
filter_control() options nfolds and mixparm. As there is some variability due to
fold selection, filterseed may also be set within filter_control(). Further
options are available for random forest selection, or as additional inputs to the
glmnet() or randomForestSRC() functions via `...'.

Filtering output is given for the run5STAR() function. For elastic net filtering,
this includes the list of candidate covariates which pass the filtering step,
details on the optimal deviance, mixing, and tuning parameters,
the elastic net beta coefficients from the optimal model, and plots of the 
deviance/parameter tuning and solution paths.

```{r filter5STAR}
cc$filteredcovs
cc$filterdetails$cvout
cc$filterdetails$beta
```

```{r filter5STARplot, fig.width=10,fig.height=5}
plot(cc$filterdetails$ENETplot)
```

### Strata Formation via CTree

5-STAR uses the CTree algorithm of Hothorn et al. (2006) to form risk-based strata
blinded to treatment assignment. This is done in two sub-steps. First, preliminary
trees are fitted on raw covariates that passed through the filtering step. This
forms preliminary strata, which are ordered by risk level using area under the Kaplan Meier curve up to the minimax time (minimum of maximum survival times over
all preliminary risk strata). These ordered strata are used as an ordered factor as
input into the CTree algorithm again to ensure no over-stratification is performed.

Many options are available for control over the strata building step via the tree.hyper option, controlled by the tree_control() function. This mainly consists of options passed through to ctree_control() from the partykit package. Defaults include minimum number of subjects/terminal node (minbucket) of max(50,5% of number of subjects), maximum tree depth of 2 (assuming no more than 4 strata), and multiplicity-adjusted significance levels for split tests of 0.1 for preliminary trees final trees (alpha). 

For our example data, 4 preliminary strata are formed via the CTree algorithm:

```{r prelimtrees5STAR, fig.width = 7, fig.height = 4}
cc$prelimstratadefn
plot(cc$prelimtree)
plot(cc$prelimbetweenstrataPlot)
```

and 3 final strata:

```{r trees5STAR, fig.width = 7, fig.height = 4}
cc$stratadefn
plot(cc$finaltree)
plot(cc$betweenstrataPlot)
```

Subject-level formed strata ids can be obtained if needed for further analysis
cc\$prelimstrataids for preliminary strata and cc\$strataids for final strata.

### Estimation within each formed strata

When measure = "HR", Cox Proportional Hazards models will be fit within each
stratum using the coxbystrata() sub function. When measure = "TR", model averaging
of AFT models will be performed using the maaftbystrata() subfunction. By default 
this model averaging is done over 3 parametric distributions, as selected through
the "distList" option: "loglogistic", "weibull", and "lognormal". 

Stratum-level summaries are provided for preliminary and final strata. Below we see the results for the final strata for our hypothetical example:

```{r step4, fig.width=16, fig.height=6}
cc$bystratasummary
print(cc$bystrataPlot)
```

In the table above, beta gives the estimand of interest on the log scale (e.g., 
log hazard ratio or log time ratio).

### Amalgamation of strata-level results

Results are amalgamated to form a final overall treatment effect estimate. The overall results for the hypothetical example is shown below.

```{r finalresults, fig.width=16, fig.height=4}
cc$res5star
print(cc$forestplot)
```
